PineAPPL:\\ EW and mixed QCD--EW corrections for PDF fits

### 2020-08-07 paper/app-pineappl.tex
Add more smaller fixes

The command-line program \texttt{pineappl} is compiled and installed using
\begin{verbatim}
 $ cargo install pineappl_cli
\end{verbatim}
This program also needs \texttt{LHAPDF} \cite{Buckley:2014ana} installed.
For usage instructions simply type \texttt{pineappl} in your shell and read the help message.

>>>>

The command-line program \texttt{pineappl} is compiled and installed using
\begin{verbatim}
 $ cargo install pineappl_cli
\end{verbatim}
This program also needs \texttt{LHAPDF} \cite{Buckley:2014ana} installed; make sure that the environment variables \texttt{PATH}, \texttt{LD\_LIBRARY\_PATH}, and \texttt{PKG\_CONFIG\_PATH} are properly set.
For usage instructions simply type \texttt{pineappl} in your shell and read the help message.

### 2020-08-07 paper/sec-conclusion.tex
Add more smaller fixes

Finally, we discussed the issue of subtracting EW corrections in experimental data, which becomes an important consideration when combining the interpolations grids with the corresponding experimental data in a PDF fit including EW and mixed QCD--EW corrections.

>>>>

Finally, we discussed the issue of subtracting EW corrections in experimental data, which becomes an important consideration when combining the interpolation grids with the corresponding experimental data in a PDF fit including EW and mixed QCD--EW corrections.

### 2020-08-07 paper/sec-introduction.tex
Add more smaller fixes

With the recent completion of Run II, the Large Hadron Collider (LHC) has 
accumulated data from an integrated luminosity of approximately 
\SI{150}{\per\femto\barn}~\cite{Mangano:2020icy}. This represents only a small fraction of
the anticipated \SI{3000}{\per\femto\barn} that will eventually be recorded in the
forthcoming twenty years of LHC operation. Nevertheless the statistical 
uncertainty of the data has already shrunk to unprecedented small values,
typically \SI{1}{\percent} or less, a fact that will allow for precision tests of the
Standard Model (SM) and for indirect searches of New Physics only if 
theoretical predictions become comparatively precise. This entails the 
computation of additional higher-order contributions to the fixed-order 
coupling-constant perturbative expansion, on the one hand, and an increasingly 
sophisticated determination of the Parton Distribution Functions (PDFs) of the 
proton~\cite{Gao:2017yyd,Ethier:2020way}, on the other hand.

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also ref.~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see e.g.\ ref.~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the electroweak (EW) and mixed QCD+EW theory has
also witnessed a comprehensive progress. Frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD--EW terms), is automated~\cite{Kallweit:2014xda,Biedermann:2017yoi,Frederix:2018nkq}. For an extensive and recent review, see ref.~\cite{Denner:2019vbn}.

@@

The systematic inclusion of EW corrections in a PDF determination entails
the solution of two separate problems. First, a problem of efficiency:
fast-interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are a fundamental ingredient
in any PDF fit, where the evaluation of the hadronic cross section needs to 
be performed a large number of times. While two default formats exist for
these grids, \textsc{APPLgrid}~\cite{Carli:2010rw} and
\textsc{fastNLO}~\cite{Kluge:2006xs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QCD+EW\@. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such (e.g.\ the $t$-channel photon-induced component in
neutral-current Drell--Yan, which is not a separate processes beyond the leading
order) or of just a part of the EW effects (e.g.\ multiple-photon
radiation from light particles in the final state of neutral- or charged-current
Drell--Yan, especially with electrons). Be that as it may, if EW effects are
systematically included in theoretical predictions, they should not be subtracted
from experimental results, otherwise they will be double counted.

>>>>

With the recent completion of Run II, the Large Hadron Collider (LHC) has 
accumulated data from an integrated luminosity of approximately 
\SI{150}{\per\femto\barn}~\cite{Mangano:2020icy}. This represents only a small fraction of
the anticipated \SI{3000}{\per\femto\barn} that will eventually be recorded in the
forthcoming twenty years of LHC operation. Nevertheless the statistical 
uncertainty of the data has already shrunk to unprecedentedly small values,
typically \SI{1}{\percent} or less, a fact that will allow for precision tests of the
Standard Model (SM) and for indirect searches of New Physics only if 
theoretical predictions become comparatively precise. This entails the 
computation of additional higher-order contributions to the fixed-order 
coupling-constant perturbative expansion, on the one hand, and an increasingly 
sophisticated determination of the Parton Distribution Functions (PDFs) of the 
proton~\cite{Gao:2017yyd,Ethier:2020way}, on the other hand.

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to a
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also ref.~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see e.g.\ ref.~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the electroweak (EW) and mixed QCD+EW theory has
also witnessed a comprehensive progress. Frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD--EW terms), is automated~\cite{Kallweit:2014xda,Biedermann:2017yoi,Frederix:2018nkq}. For an extensive and recent review, see ref.~\cite{Denner:2019vbn}.

@@

The systematic inclusion of EW corrections in a PDF determination entails
the solution of two separate problems. First, a problem of efficiency:
fast-interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are a fundamental ingredient
in any PDF fit, where the evaluation of the hadronic cross section needs to 
be performed a large number of times. While two default formats exist for
these grids, \textsc{APPLgrid}~\cite{Carli:2010rw} and
\textsc{fastNLO}~\cite{Kluge:2006xs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QCD+EW\@. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such (e.g.\ the $t$-channel photon-induced component in
neutral-current Drell--Yan, which is not a separate processes beyond leading
order) or of just a part of the EW effects (e.g.\ multiple-photon
radiation from light particles in the final state of neutral- or charged-current
Drell--Yan, especially with electrons). Be that as it may, if EW effects are
systematically included in theoretical predictions, they should not be subtracted
from experimental results, otherwise they will be double counted.

### 2020-08-07 paper/sec-pineappl.tex
Add more smaller fixes

In general, we define leading order (LO) as the set of contributions for all possible initial states $a b$ that lead to the same final state $X$, for which the sum of the coupling exponents in eq.~\eqref{eq:expansion} is smallest, i.e.\ $k + l = p$, where $p = \min (k+l)$.
This number is process dependent and usually determined by the number of external particles.
For many processes there is only one LO, but when a process has multiple quark lines, colourless (photons, \dots) and coloured particles (gluons, \ldots) can be exchanged between them, making it possible to have more than one LO.
An important example at the LHC is dijet production, which has three different LOs: $\mathcal{O} (\alpha^2)$, $\mathcal{O} (\alphas \alpha)$, and $\mathcal{O} (\alpha^2)$.
Each LO receives a higher-order correction with an additional power of $\alphas$ or $\alpha$, which in general leads to at least two next-to-leading order (NLO) corrections.
The correction with the largest power in $\alphas$ can be unambiguously called \enquote{the} QCD correction, and the one with the largest power in $\alpha$ \enquote{the} EW correction.
All remaining corrections are of mixed type, meaning that they, in general, cannot be attributed to either one of strong or electroweak origin.
%Sometimes there are higher orders that cannot directly be understood as a correction to a LO, e.g.\ $\mathrm{p} \mathrm{p} \to \ell \bar{\ell} + \mathrm{jet} + X$, which has one LO, $\mathcal{O} (\alphas \alpha^2)$, but three NLOs, one of which is $\mathcal{O} (\alpha^4)$~TODO.

Due to the typical sizes of the couplings $\alphas^2 \sim \alpha$, it is naively expected that within the same order, i.e.\ for fixed $k + l$, terms with larger powers $\alphas^k$ dominate over those with smaller powers.
In practice, however, this naive expectation is not always true due to dynamic effects.
Some examples are vector-boson scattering processes~\cite{Biedermann:2017bss,Denner:2019tmn}, top-pair production with a W boson and four-top production~\cite{Frederix:2017wme}, and Higgs production with a bottom-pair~\cite{Pagani:2020rsg}.

@@

Using 4-tuples has the clear advantage of being very easy to implement and test.
Furthermore, they reproduce the exact numerical value that is also calculated by the MC integrator.
However, the price one has to pay is the size of the 4-tuples.
To give a concrete example, NLO QCD+EW Drell--Yan lepton-pair production (see section~\ref{sec:pineappl-example} and section~\ref{subsec:numerical_results}) needs \SI{159}{\giga\byte} of storage for a target precision of \SI{1}{\percent} of the integrated cross section.
While this is an acceptable size, increasing the precision by an order of magnitude would require roughly 100 times the size, due to the Monte Carlo convergence that goes as $1/\sqrt{N}$ with $N$ being the number of 4-tuples.
With increasing size also the speed of the convolution degrades, because it basically becomes bound by the speed with which the 4-tuples can be read from disk.
However, due to the uncompressed nature of this representation it can serve as an intermediate format to develop and quickly cross check more space-efficient representations, which we will discuss next.

@@

This method first maps $(x_1, x_2, Q^2) \mapsto (y_1, y_2, \tau)$, with
\begin{equation}
y(x) = 5 (1-x) - \log x \text{,} \qquad \tau (Q^2) = \log \log \frac{Q^2}{(\SI{0.25}{\giga\electronvolt})^2} \text{.}
\label{eq:maps}
\end{equation}
The function $y(x)$ maps events with large $x$ effectively linearly and small $x$ effectively logarithmically onto $y$.
This reflects our knowledge of PDFs, which behave differently in those regions, and thereby increases the precision of the interpolation.
For the convolution of a grid with a PDF set also the inverse functions are needed, which are
\begin{equation}
x(y) = \frac{1}{5} \operatorname{W}_0 (5 \exp (5-y)) \text{,} \qquad Q^2 (\tau) = (\SI{0.25}{\giga\electronvolt})^2 \exp (\exp (\tau)) \text{,}
\end{equation}
where $\operatorname{W}_0 (x)$ is (the principle branch of) the Lambert W function or product logarithm, which satisfies the relation $\operatorname{W} (x) \exp (\operatorname{W} (x)) = x$.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we assume we have chosen a static scale, so that we do not need to interpolate in $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them\footnote{The technical term for this interpolation problem is sometimes called \emph{subtabulation}.}.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u_1(y_1(x_1)),u_2(y_2(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} = a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting the smallest index in the square, $k_i$ and $k_j$.
Finally, we remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part gives the location between the nearest grid points.

>>>>

In general, we define leading order (LO) as the set of contributions for all possible initial states $a b$ that lead to the same final state $X$, for which the sum of the coupling exponents in eq.~\eqref{eq:expansion} is smallest, i.e.\ $k + l = p$, where $p = \min (k+l)$.
This number is process dependent and usually determined by the number of external particles.
For many processes there is only one LO, but when a process has multiple quark lines, colourless (photons, \dots) and coloured particles (gluons, \ldots) can be exchanged between them, making it possible to have more than one LO.
An important example at the LHC is (on-shell) top-pair production, which has three different LOs: $\mathcal{O} (\alpha^2)$, $\mathcal{O} (\alphas \alpha)$, and $\mathcal{O} (\alpha^2)$.
Each LO receives a higher-order correction with an additional power of $\alphas$ or $\alpha$, which in general leads to at least two next-to-leading order (NLO) corrections.
The correction with the largest power in $\alphas$ can be unambiguously called \enquote{the} QCD correction, and the one with the largest power in $\alpha$ \enquote{the} EW correction.
All remaining corrections are of mixed type, meaning that they, in general, cannot be attributed to either one of strong or electroweak origin.
%Sometimes there are higher orders that cannot directly be understood as a correction to a LO, e.g.\ $\mathrm{p} \mathrm{p} \to \ell \bar{\ell} + \mathrm{jet} + X$, which has one LO, $\mathcal{O} (\alphas \alpha^2)$, but three NLOs, one of which is $\mathcal{O} (\alpha^4)$~TODO.

Due to the typical sizes of the couplings $\alphas^2 \sim \alpha$, it is naively expected that within the same order, i.e.\ for fixed $k + l$, terms with larger powers $\alphas^k$ dominate over those with smaller powers (and larger powers of $\alpha^l$).
In practice, however, this naive expectation is not always true due to dynamic effects.
Some examples are vector-boson scattering processes~\cite{Biedermann:2017bss,Denner:2019tmn}, top-pair production with a W boson and four-top production~\cite{Frederix:2017wme}, and Higgs production with a bottom-pair~\cite{Pagani:2020rsg}.

@@

Using 4-tuples has the clear advantage of being very easy to implement and test.
Furthermore, they reproduce the exact numerical value that is also calculated by the MC integrator.
However, the price one has to pay is the size of the 4-tuples.
To give a concrete example, NLO QCD+EW Drell--Yan lepton-pair production (see section~\ref{sec:pineappl-example} and section~\ref{sec:dy-lepton-pair-production}) needs \SI{159}{\mega\byte} of storage for a target precision of \SI{1}{\percent} of the integrated cross section.
While this is an acceptable size, increasing the precision by an order of magnitude would require roughly 100 times the size, due to the Monte Carlo convergence that goes as $1/\sqrt{N}$ with $N$ being the number of 4-tuples.
With increasing size also the speed of the convolution degrades, because it basically becomes bound by the speed with which the 4-tuples can be read from disk.
However, due to the uncompressed nature of this representation it can serve as an intermediate format to develop and quickly cross check more space-efficient representations, which we will discuss next.

@@

This method first maps $(x_1, x_2, Q^2) \mapsto (y_1, y_2, \tau)$, with
\begin{equation}
y(x) = 5 (1-x) - \log x \text{,} \qquad \tau (Q^2) = \log \log \frac{Q^2}{(\SI{0.25}{\giga\electronvolt})^2} \text{.}
\label{eq:maps}
\end{equation}
The function $y(x)$ maps events with large $x$ effectively linearly and small $x$ effectively logarithmically onto $y$.
This reflects our knowledge of PDFs, which behave differently in those two regions, and thereby increases the precision of the interpolation.
For the convolution of a grid with a PDF set also the inverse functions are needed, which are
\begin{equation}
x(y) = \frac{1}{5} \operatorname{W}_0 (5 \exp (5-y)) \text{,} \qquad Q^2 (\tau) = (\SI{0.25}{\giga\electronvolt})^2 \exp (\exp (\tau)) \text{,}
\end{equation}
where $\operatorname{W}_0 (x)$ is (the principle branch of) the Lambert W function or product logarithm, which satisfies the relation $\operatorname{W} (x) \exp (\operatorname{W} (x)) = x$.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we have chosen a static scale, so that we do not need to interpolate in $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them\footnote{The technical term for this interpolation problem is sometimes called \emph{subtabulation}.}.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u_1(y_1(x_1)),u_2(y_2(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} = a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting the smallest index in the square, $k_i$ and $k_j$.
Finally, we remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part gives the location between the nearest grid points.

% TODO: explain how convolutions are performed

### 2020-08-07 paper/sec-results.tex
Fix reference
Add more smaller fixes

In this section we demonstrate the capabilities of \textsc{PineAPPL} by
computing fast-interpolation grids, accurate to NLO QCD and NLO QCD+EW,
for a representative set of processes in which EW corrections may have a
sizeable effect on the accuracy of the theoretical predictions.
In order to consider some realistic kinematics for these
processes, we resort to measurements commonly devised for inclusion in PDF
fits. Our aim is twofold. First, we want to validate the results
obtained with \textsc{PineAPPL}; second, we want to assess the
impact of the EW corrections for usual experimental setups. We start by describing the
setup employed for the computation, then we discuss the features of each process, along with
the corresponding results.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Drell--Yan lepton pair production.}

>>>>

In this section we demonstrate the capabilities of \textsc{PineAPPL} by
computing fast-interpolation grids, accurate to NLO QCD and NLO QCD+EW,
for a representative set of processes in which EW corrections may have a
sizeable effect on the accuracy of the theoretical predictions.
In order to consider realistic kinematics for these
processes, we rely on the measurements that are commonly included in PDF
fits. Our aim is twofold. First, we want to validate the results
obtained with \textsc{PineAPPL}; second, we want to assess the
impact of the EW corrections for the specific experimental setups. We start by describing the
setup employed for the computation, then we discuss the features of each process, along with
the corresponding results.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Drell--Yan lepton pair production.}
\label{sec:dy-lepton-pair-production}
